Lesson 3
========================================================

### Pseudo-Facebook User Data
Notes:

```{r Pseudo-Facebook User Data}
library(ggplot2)

setwd("C:/Users/Christopher/Desktop/Nanodegree/P4-EDA/eda-course-materials/lesson3")

pfd <- read.csv("pseudo_facebook.tsv", sep = "\t")

```
***

### Histogram of Users' Birthdays
Notes:

```{r Histogram of Users\' Birthdays}
qplot(data = pfd, x = dob_day) + 
  scale_x_continuous(breaks=1:31)
```

***

### Faceting
Notes:

```{r Faceting}
ggplot(data = pfd, aes(x = dob_day)) +
  geom_histogram(binwidth = 1) +
  scale_x_continuous(breaks = 1:31) +
  facet_wrap(~dob_month)
```

#### What code would you enter to create a histogram of friend counts?

```{r Friend Count}
qplot(data=pfd, x=friend_count)
```

### Limiting the Axes
Notes:

```{r Limiting the Axes}
qplot(data=pfd, x=friend_count, xlim = c(0,1200))

qplot(data=pfd, x = friend_count) + scale_x_continuous(limits = c(0,1200))
```

***

### Adjusting the Bin Width
Notes:

### Faceting Friend Count
```{r Faceting Friend Count}
# What code would you add to create a facet the histogram by gender?
# Add it to the code below.
ggplot(aes(x = friend_count), data = pfd) +
  geom_histogram(binwidth = 25) +
  scale_x_continuous(limits = c(0, 1000), breaks = seq(0, 1000, 50)) + facet_wrap(facets=~gender)
```
```{r}
qplot(x = friend_count, data = subset(pfd, !is.na(gender)), xlim=c(0,1000)) +
  facet_wrap(~gender)
```


***

### Omitting NA Values
Notes:

```{r Omitting NA Values}
qplot(data=na.omit(pfd), x=friend_count, xlim=c(0,1000))
```

***

### Statistics 'by' Gender
Notes:

```{r Statistics \'by\' Gender}
table(pfd$gender)
by(pfd$friend_count, pfd$gender, summary)
```

***

### Tenure
Notes:

```{r Tenure}
ggplot(aes(x = tenure/365), data = pfd) +
   geom_histogram(binwidth = .25, color = I('black'), fill = I('#099DD9')) +scale_x_continuous(breaks=seq(0,10,.5)) +xlab("Years using Facebook") + ylab("Number of Users in Sample")
```

***

### User Ages
Notes:

```{r User Ages}
ggplot(data = pfd, aes(x=age)) + geom_histogram(binwidth=1, color = I('black'), fill = I('#099DD9')) +
  scale_x_continuous(breaks=seq(0,130,10),limits = c(min(pfd$age),max(pfd$age))) + 
  xlab("User Age")+ylab("Count")
  
```


### Add a Scaling Layer
Notes:

```{r Add a Scaling Layer}
library(gridExtra)
p1 = ggplot(data = pfd, aes(x=friend_count)) + geom_histogram() +
 xlab("Friend Count")+ylab("Count")

p2 = p1 + scale_x_log10() + xlab("Friend Count log10") + ylab("Count")

p3 = p1 + scale_x_sqrt() + xlab("Sqrt Friend Count") + ylab("Count")

p = grid.arrange(p1,p2,p3)

```

***


### Frequency Polygons

```{r Frequency Polygons}
ggplot(aes(x = friend_count, y = ..density../sum(..density..)), data = subset(pfd, !is.na(gender))) +
  geom_freqpoly(aes(color = gender), binwidth=10) +
  scale_x_continuous(limits = c(0, 1000), breaks = seq(0, 1000, 50)) +
  xlab('Friend Count') +
  ylab('Percentage of users with that friend count')
```

***

### Likes on the Web
Notes:

```{r Likes on the Web}
ggplot(aes(x = www_likes, y = ..density../sum(..density..)), data = subset(pfd, !is.na(gender))) +
  geom_freqpoly(aes(color = gender), binwidth=1) +
  scale_x_log10() + xlab('Webpage Likes') + ylab('Percentage of users')

by(pfd$www_likes, pfd$gender, sum)
```


***

### Box Plots
Notes:

```{r Box Plots}
ggplot(data=subset(pfd,!is.na(gender)), aes(x = gender, y = friend_count)) + geom_boxplot() + scale_y_continuous(limits = c(0, 1000))


```

#### Adjust the code to focus on users who have friend counts between 0 and 1000.

```{r}
ggplot(data=subset(pfd,!is.na(gender)), aes(x = gender, y = friend_count)) + geom_boxplot() + coord_cartesian(ylim = c(0, 1000))
```
Using ylim or y_scale_continuous removes data points from box plot calculations; coord_cartesian does not.
***


### Getting Logical
Notes:

```{r Getting Logical}
mobile_check_in <- NA
pfd$mobile_check_in <- pfd$mobile_likes > 0
sum(pfd$mobile_check_in)/length(pfd$mobile_check_in)
```

```{r Another Method}
pfd$mobile_check_in <- ifelse(pfd$mobile_likes > 0, 1, 0)
pfd$mobile_check_in <- factor(pfd$mobile_check_in)
sum(pfd$mobile_check_in == 1)/length(pfd$mobile_check_in)
```


Response:

***

### Analyzing One Variable
Reflection:

***

Click **KnitHTML** to see all of your hard work and to have an html
page of this lesson, your answers, and your notes!