Lesson 4 ======================================================== 
```{r Pseudo-Facebook User Data} 
knitr::opts_chunk$set(echo = T, fig.width = 7, fig.align = "center", out.width = "80%")
library(ggplot2) 
library(dplyr, tidyr)

setwd("C:/Users/Christopher/Desktop/Nanodegree/P4-EDA/eda-course-materials/")

pfd <- read.csv("pseudo_facebook.tsv", sep = "\t")
``` 

***

### Scatterplots and Perceived Audience Size Notes:

***

### Scatterplots Notes:



```{r Scatterplots} 
ggplot(aes(x = age, y = friend_count), data = pfd) + 
  geom_jitter(alpha=1/20) + xlim(c(13,90))
```

***

#### What are some things that you notice right away? Response:

***

### ggplot Syntax Notes:

```{r ggplot Syntax}
ggplot(aes(x = age, y = friend_count), data = pfd) + 
  geom_jitter(alpha=1/20) + xlim(c(13,90))
```

***

### Overplotting Notes:

```{r Overplotting}
ggplot(aes(x = age, y = friend_count), data = pfd) +
  geom_jitter(alpha=1/20) + xlim(c(13,90))
```

#### What do you notice in the plot? Response:

***

### Coord_trans() Notes:

```{r Coord_trans()}
ggplot(aes(x = age, y = friend_count), data = pfd) +
  geom_point(alpha=1/20) + xlim(c(13,90)) +
  coord_trans(y="sqrt")
```

#### Look up the documentation for coord_trans() and add a layer to the plot
that transforms friend_count using the square root function. Create your plot!

```{r}
ggplot(aes(x = age, y = friend_count), data = pfd) +
  geom_point(alpha=1/20) + xlim(c(13,90)) +
  coord_trans(y="sqrt")
```

#### What do you notice? Jitter doesn't work because it "jitters" the data
points, inserting noise between points to spread them out a bit. For our
zero-valued friend counts, this can lead to negative values, and the sqrt
coord_trans then returns an error. If R were better-written, the error would
point out that sqrt is being passed negative numbers, but it doesn't. ***

### Alpha and Jitter Notes:

```{r Alpha and Jitter}
ggplot(aes(x = age, y = friend_count), data = pfd) + 
geom_jitter(position=position_jitter(h=0), alpha=1/20) + xlim(c(13,90)) + 
coord_trans(y="sqrt")
```

*** 

```{r Friends_initiated versus Age}
ggplot(aes(x= age, y=friendships_initiated), data=pfd) + 
  geom_point(alpha=1/20, position=position_jitter(h=0)) + 
  xlim(c(13,90)) + coord_trans(y='sqrt')
``` 

### Overplotting and Domain Knowledge Notes:

***

### Conditional Means Notes:

```{r Conditional Means} 
agegroups = group_by(pfd, age) 
pfd.fc_by_age = summarize(agegroups, 
                          friend_count_mean = mean(friend_count),
                          friend_count_median = median(friend_count),
                          n=n()) 
pfd.fc_by_age = arrange(pfd.fc_by_age, age) 
```

```{r Conditional Means thru Chaining}
pfd.fc_by_age <- group_by(pfd, age) %>%
  summarize(friend_count_mean = mean(friend_count),
            friend_count_median = median(friend_count),
            n=n()) %>% arrange(age)
```

Create your plot!

```{r Conditional Means Plot} 
ggplot(aes(x = age, y = friend_count_mean), data = pfd.fc_by_age) +
  geom_line()
```

***

### Overlaying Summaries with Raw Data Notes:

```{r Overlaying Summaries with Raw Data}
ggplot(aes(x = age, y = friend_count), data = pfd) +
  geom_point(position = position_jitter(h=0),
             alpha=1/20,
             color='orange') +
  coord_cartesian(xlim = c(13, 90)) +
  geom_line(stat='summary', fun.y=mean) +
  geom_line(stat='summary', fun.y=median, linetype=2) +
  geom_line(stat="summary", fun.y=quantile,
            fun.args = list(probs = .9),
            color="blue" ) +
  geom_line(stat="summary", fun.y=quantile,
            fun.args = list(probs = .1), linetype=2, color="blue")
```

#### What are some of your observations of the plot? Response:

***

### Moira: Histogram Summary and Scatterplot See the Instructor Notes of this
video to download Moira's paper on perceived audience size and to see the final
plot.

Notes:

***

### Correlation Notes:

```{r Correlation using the with function}
with(pfd, cor.test(age, friend_count, method='pearson'))
```

Look up the documentation for the cor.test function.

What's the correlation between age and friend count? Round to three decimal
places. Response:

***

### Correlation on Subsets Notes:

```{r Correlation on Subsets}

```

***

### Correlation Methods Notes: cor.test defaults to pearson's ***

## Create Scatterplots Notes:

```{r}
ggplot(data=pfd, aes(x=www_likes_received, y=likes_received)) +
  geom_point() + 
  xlim(0, quantile(pfd$likes_received, 0.95)) +
  ylim(0, quantile(pfd$www_likes_received, 0.95)) +
  geom_smooth(method = 'lm', color = 'red')
```

***

### Strong Correlations Notes:

```{r Strong Correlations}
with(subset(pfd, age <= 70),
     cor.test(www_likes_received, likes_received))
```

What's the correlation betwen the two variables? Include the top 5% of values
for the variable in the calculation and round to 3 decimal places.

```{r Correlation Calcuation}
with(pfd, cor.test(www_likes_received, likes_received))
```

Response:

***

### Moira on Correlation Notes:

***

### More Caution with Correlation Notes:

```{r More Caution With Correlation}
library(alr3) 
data(Mitchell)
```

Create your plot!

```{r Temp vs Month}
ggplot(data=Mitchell, aes(x=Month, y=Temp)) + geom_point()
```

***

### Noisy Scatterplots a. Take a guess for the correlation coefficient for the
scatterplot.

b. What is the actual correlation of the two variables? (Round to the
thousandths place)

```{r Noisy Scatterplots}
cor.test(Mitchell$Month, Mitchell$Temp)
```

***

### Making Sense of Data Notes:

```{r Making Sense of Data}
ggplot(data=Mitchell, aes(x=Month%%12, y=Temp)) + geom_point() +
  scale_x_continuous(breaks=seq(0,203,12))

Mitchell$stdmonth <- Mitchell$Month%%12
by_month <- group_by(Mitchell, stdmonth)
mthlymeans <- by_month %>% summarize(mean_temp = mean(Temp))
```

***

### A New Perspective

What do you notice? Response:

Watch the solution video and check out the Instructor Notes! Notes:

***

### Understanding Noise: Age to Age Months Notes:

```{r Understanding Noise: Age to Age Months}
pfd$age_with_months <- pfd$age + (1-pfd$dob_month/12)
head(pfd$age_with_months,10)
```

***

### Age with Months Means

```{r Age with Months Means}
pfd.fc_by_age_months <-  pfd %>%
  group_by(age_with_months) %>%
  summarize(friend_count_mean = mean(friend_count),
            friend_count_median= median(friend_count),
            n = n()) %>%
  arrange(age_with_months)
```

Programming Assignment 
```{r Programming Assignment}

```

***

### Noise in Conditional Means

```{r Noise in Conditional Means}
library(gridExtra)

p1 = ggplot(data=pfd.fc_by_age_months, aes(x=age_with_months, y=friend_count_mean)) + 
  geom_line() +
  geom_smooth() +
  xlim(c(13,71))

p2 = ggplot(data=pfd, aes(x=round(age/5)*5, y=friend_count)) + 
  geom_line(stat="summary", fun.y=mean) +
  geom_smooth() +
  xlim(c(13,71))
grid.arrange(p1,p2)
```

***

### Smoothing Conditional Means Notes:

```{r Smoothing Conditional Means}

```

***

### Which Plot to Choose? Notes:

***

### Analyzing Two Variables Reflection:

***

Click **KnitHTML** to see all of your hard work and to have an html page of this
lesson, your answers, and your notes!