Lesson 5
========================================================
```{r Setup}
library(ggplot2)
library(tidyr)
library(dplyr)
setwd("C:/Users/Christopher/Desktop/Nanodegree/P4-EDA/eda-course-materials")

pf <- read.csv("pseudo_facebook.tsv", header = T, sep = "\t")
```

### Multivariate Data
Notes:

***

### Moira Perceived Audience Size Colored by Age
Notes:

***

### Third Qualitative Variable
Notes:

```{r Third Qualitative Variable}
ggplot(aes(x = gender, y = age),
       data = subset(pf, !is.na(gender))) + 
  geom_boxplot() + stat_summary(fun.y = mean, geom = 'point', shape=4)

ggplot(aes(x = age, y = friend_count),
       data = subset(pf, !is.na(gender))) + 
  geom_line(aes(color=gender), stat = "summary", fun.y = median)

```

***

### Plotting Conditional Summaries
Notes:

```{r Plotting Conditional Summaries}
pf.fc_by_age_gender <- dplyr::group_by(pf, age, gender) %>% 
  dplyr::summarize(mean_friend_count = mean(friend_count), 
            median_friend_count = median(friend_count),
            n = n())

ggplot(aes(x = age, y = median_friend_count),
       data = subset(pf.fc_by_age_gender, !is.na(gender))) + 
  geom_line(aes(color=gender))
```

***

### Thinking in Ratios
Notes:

***

### Wide and Long Format
Notes:
If long includes repeated and discrete variables like age or gender, it is useful at times to create new dfs that aggregate these data with calculated parameters like median or mean. Categorical variables like gender are simply counted up in the wide format.
***

### Reshaping Data
Notes:

```{r Reshpaing with reshape2}
library(reshape2)

pf.fc_by_age_gender.wide <- dcast(data = pf.fc_by_age_gender, 
                                  age ~ gender,
                                  value.var = "median_friend_count")
pf.fc_by_age_gender.wide$ratio <- 
  pf.fc_by_age_gender.wide$male/pf.fc_by_age_gender.wide$female
```


```{r Reshaping with Tidyr and dplyr}
pf.fc_by_age_gender.wide <- subset(
  pf.fc_by_age_gender[c('age', 'gender', 'median_friend_count')], 
  !is.na(gender)) %>% 
  spread(gender, median_friend_count) %>% 
  mutate(ratio = male / female)
```


***

### Ratio Plot
Notes:
The linetype parameter can take the values 0-6:
0 = blank, 1 = solid, 2 = dashed
3 = dotted, 4 = dotdash, 5 = longdash
6 = twodash

```{r Ratio Plot}
ggplot(data = pf.fc_by_age_gender.wide, aes(x=age, y=female/male)) +
  geom_line() +
  geom_hline(yintercept = 1, linetype = 2, alpha=0.5)
```

***

### Third Quantitative Variable
Notes:

```{r Third Quantitative Variable}
pf$year_joined <- floor(2014-pf$tenure/365)
```

***

### Cut a Variable
Notes: Pretty pandas-like

```{r Cut a Variable}
pf$year_joined.bucket <- cut(pf$year_joined, c(2004, 2009, 2011, 2012, 2014), labels = c("2004-2009", "2009-2011", "2011-2012", "2012-2014"))

table(pf$year_joined.bucket, useNA = "ifany")
```

***

### Plotting it All Together
Notes:

```{r Plotting it All Together}
ggplot(data = subset(pf, !is.na(gender)), 
       aes(x=age, y=friend_count)) +
  geom_line(aes(color=year_joined.bucket), stat="summary",
            fun.y=median)
```

***

### Plot the Grand Mean
Notes:

```{r Plot the Grand Mean}
ggplot(data = subset(pf, !is.na(gender) | !is.na(year_joined.bucket)), 
       aes(x=age, y=friend_count)) +
  geom_line(aes(color=year_joined.bucket), stat="summary",
            fun.y=mean) +
  geom_line(stat="summary", fun.y=mean, linetype = 2)
```

***

### Friending Rate
Notes:

```{r Friending Rate}
with(subset(pf, tenure >0), summary(friend_count/tenure))
```

***

### Friendships Initiated
Notes:

What is the median friend rate?

What is the maximum friend rate?

```{r Friendships Initiated}
ggplot(data=subset(pf, tenure > 0), 
       aes(x = tenure, 
           y = friendships_initiated/tenure)) + geom_line(aes(color=year_joined.bucket), stat="summary", fun.y=mean) +
  geom_smooth()
```

***

### Bias-Variance Tradeoff Revisited
Notes:

```{r Bias-Variance Tradeoff Revisited}

ggplot(aes(x = tenure, y = friendships_initiated / tenure),
       data = subset(pf, tenure >= 1)) +
  geom_line(aes(color = year_joined.bucket),
            stat = 'summary',
            fun.y = mean)

ggplot(aes(x = 7 * round(tenure / 7), y = friendships_initiated / tenure),
       data = subset(pf, tenure > 0)) +
  geom_line(aes(color = year_joined.bucket),
            stat = "summary",
            fun.y = mean)

ggplot(aes(x = 30 * round(tenure / 30), y = friendships_initiated / tenure),
       data = subset(pf, tenure > 0)) +
  geom_line(aes(color = year_joined.bucket),
            stat = "summary",
            fun.y = mean)

ggplot(aes(x = 90 * round(tenure / 90), y = friendships_initiated / tenure),
       data = subset(pf, tenure > 0)) +
  geom_line(aes(color = year_joined.bucket),
            stat = "summary",
            fun.y = mean)

```

***

### Sean's NFL Fan Sentiment Study
Notes:

***

### Introducing the Yogurt Data Set
Notes:

***

### Histograms Revisited
Notes:

```{r Histograms Revisited}
yo <- read.csv("yogurt.csv", header = T)
yo$id <- factor(yo$id)
str(yo)
ggplot(data=yo, aes(x=price)) + geom_histogram(fill= "blue")
```

***

### Number of Purchases
Notes:

```{r Number of Purchases}
yo <- transform(yo, all.purchases = strawberry + blueberry + pina.colada + plain + mixed.berry)
```

***

### Prices over Time
Notes:

```{r Prices over Time}
ggplot(data=yo, aes(x=time, y=price)) +
  geom_jitter(alpha=.25, shape = 21, fill="blue")
```

***

### Sampling Observations
Notes:
With so many repeat measurements it is useful to take a random subset of the data (in this case purchases for a household), to be able to more closely examine variation across data points (households).
***

### Looking at Samples of Households

```{r Looking at Sample of Households}
set.seed(4230)
samp.ids <- sample(levels(yo$id), 16)

ggplot(data=subset(yo, id %in% samp.ids), 
       aes(x=time, y=price)) + 
  facet_wrap(~id) +
  geom_line() +
  geom_point(aes(size = all.purchases), pch=1)
```

```{r Another sampling}
set.seed(201)

samp.ids <- sample(levels(yo$id), 16)

ggplot(data=subset(yo, id %in% samp.ids), aes(x=time)) +
  facet_wrap(~id) +
  geom_point(aes(y=strawberry), color="red") +
  geom_point(aes(y=blueberry), color="blue")
```

***

### The Limits of Cross Sectional Data
Notes: As in the pf data, we have no time-series data, making it "cross sectional", or a snapshot of a larger data generating process. We can't make across-time comparisons.

***

### Many Variables
Notes:

***

### Scatterplot Matrix
Notes:

```{r}
library(GGally)
theme_set(theme_minimal(20))

set.seed(1836)
pf.subset <- pf[,c(2:15)]
names(pf.subset)
ggpairs(pf.subset[sample.int(nrow(pf.subset),1000), ])
```


***

### Even More Variables
Notes:

***

### Heat Maps
Notes:

```{r}
nci <- read.table("nci.tsv")
colnames(nci) <- c(1:64)
```

```{r}
nci.long.samp <- melt(as.matrix(nci[1:200,]))
names(nci.long.samp) <- c("gene", "case", "value")
head(nci.long.samp)

ggplot(aes(y = gene, x = case, fill = value),
  data = nci.long.samp) +
  geom_tile() +
  scale_fill_gradientn(colours = colorRampPalette(c("blue", "red"))(100))
```


***

### Analyzing Three of More Variables
Reflection:

***

Click **KnitHTML** to see all of your hard work and to have an html
page of this lesson, your answers, and your notes!

